---
title: "Stacked Bars Connected by Area Chart"
author: "Fiona Lees"
date: 2022-06-25
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      out.width = "100%",
                      fig.align = "center")

```


### Objective
To work out how to create a chart similar to one [published by the Financial Times](https://twitter.com/jburnmurdoch/status/1525766121776943107), by John Burn-Murdoch. The chart contains two stacked-bars, connected by an area chart. 


### Load Packages

```{r packages}

library(showtext)
library(tidyverse)
library(ggrepel)

```

### Import Data
 
```{r import_data}

my_data <- readxl::read_excel("data/data_for_chart.xlsx")

```

```{r data_view}

view(my_data)
glimpse(my_data)

```


### Explore Options 

Create stacked bar chart.  
 
```{r explore_1}

my_data %>% 
  ggplot(aes(x = year, y = value, fill = group)) + 
  geom_bar(position = "stack", stat = "identity", width = 1)

```

Create area chart.  

```{r explore_2}

my_data %>% 
  ggplot(aes(x = year, y = value, fill = group)) + 
  geom_area()

```

Put the stacked bars and the area chart together. Adjust the start and end points of the area chart so that it meets the edges of the stacked bars.  

```{r explore_3}

my_data %>% 
  ggplot(aes(x = year, y = value, fill = group)) + 
  # stacked bars
  geom_bar(position = "stack", stat = "identity", width = 1) +
  # area chart
  geom_area(
    aes(x = ifelse(year == 2012, year + 0.5, year - 0.5)), # width adjustment
    alpha = 0.5 # add transparency
  )

```


### Visualise Data

Build the final visualisation.  

```{r fonts}

# Load fonts
showtext_auto(enable = TRUE)
font_add_google("Lato")
font_add_google("Roboto Slab")

```

```{r colours}

# Define colour palette  
my_pal <- c("#e6005c", "grey80", "#008080")
my_pal_2 <- c("#e6005c", "grey40", "#008080")

```

```{r final_chart, fig.width = 5, fig.asp = 1, fig.alt = ""}

# Create final chart
p_final <- my_data %>% 
  ggplot(aes(x = year, y = value, fill = group)) + 
  # Stacked bars
  geom_bar(position = "stack", stat = "identity", width = 2) +
  # Area chart
  geom_area(
    aes(x = ifelse(year == 2012, year + 1, year - 1)),
    alpha = 0.4
    ) +
  # Percentage labels 
  geom_text(
    aes(label = scales::percent(value)),
    position = position_stack(vjust = 0.5),
    size = 3.88,
    colour = c("white", "black", "white", "white", "black", "white")
    ) +
  # Group labels
  geom_text_repel(
    data = filter(my_data, year == 2012),
    aes(label = str_c("Group ", group)),
    position = position_stack(vjust = 0.5),
    direction = "x",
    xlim = c(2007, 2010.5),
    segment.colour = NA,
    size = 3.88,
    fontface = "bold",
    colour = my_pal_2
    ) +
  # Colour
  scale_fill_manual(values = my_pal) +
  # Scales
  scale_x_continuous(limits = c(2007, 2023), breaks = (c(2012, 2022)), position = "top") +
  scale_y_continuous(expand = expansion(add = 0)) +
  # Styling
  theme_minimal() +
  theme(
    text = element_text(family = "Lato", colour = "grey40", size = 12),
    plot.title = element_text(family = "Roboto Slab", colour = "black", size = 16, 
                              face = "bold", margin = margin(b = 5), hjust = 0),
    plot.subtitle = element_text(size = 12, hjust = 0),
    plot.caption = element_text(size = 8, margin = margin(t = 20), hjust = 0),
    plot.background = element_rect(colour = "white", fill = "white"),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 12, colour = "black"),
    panel.grid = element_blank(),
    legend.position = "none",
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
    ) +
  # Titles
  labs(
    title = "Title goes here",
    subtitle = "Subtitle goes here",
    x = "", 
    y = "",
    caption = "Visualisation: Fiona Lees (@Fi_Lees) | Inspired by a FT Graphic by John Burn-Murdoch"
    )

p_final 

```

Save the plot as an image.  

```{r save_image}

# Tell showtext to use DPI of 320 so text size is correct when chart is saved as an image
showtext_opts(dpi = 320)

# Save plot as image
ggsave("bars_connected_by_area.png", p_final, width = 5, height = 5, units = "in", dpi = 320)

# Turn off showtext
showtext_auto(FALSE)

```


### Session Information
```{r session_info}

sessionInfo()

```

